Circular Distributional Trees and Forests with an Application to Wind Direction Forecasting

### 2019-11-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revised abstract, small corrections in description/caption of Figure 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2888 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 365d174d6d1ff9c8426d47689da099be6dc29348

\Abstract{
As circular data can be found in many different areas, the development of
appropriate statistical models accounting, i.a., for periodicity has become a
challenging and important research field. For probabilistic forecasting
typically the von Mises distribution is employed in an additive model
framework. Combining their underlying distributional setup with tree-based
estimation methods, circular distributional trees will be presented. The resulting
distributional trees are easy to interpret, can detect non-additive effects,
and select covariates and their interactions automatically. By combining
ensembles of circular trees a stabilized circular forest can build, which
automatically regularizes the estimation and, hence, allows for modeling smooth effects.

To validate the proposed circular trees and forests, a case study on
probabilistic forecasting of wind direction will be shown for two airports in
Austria. In addition to the tree-based models, alternative commonly used
approaches will be employed to compare their predictive performances to the
proposed circular trees and forest.

@@

\begin{figure}[t]
%\centering
%\hspace{-1cm}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \vspace{-1em}
  \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \includegraphics[width=1.2\linewidth]{density_circular.pdf}
\end{minipage}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\caption{Illustration of a distributional model fitted to circular data by 
  maximum likelihood-estimation employing the von Mises distribution. 
  Left panel: linear illustration. Right panel: circular illustration.
  In both panels the empirical histogram (gray) and fitted density (red line) are depicted along 
  with the estimated location parameter (red hand).
  }
  \label{fig:densities} 
  %\includegraphics[width=4cm]{density_vM.pdf}
\end{figure}
\fixme{update figure 2d to 3d, using data of example tree}

Figure~\ref{fig:densities} illustrates a distributional model fitted to circular data
employing the von Mises distribution.
% on observed wind direction data from the airport of Innsbruck.
This example shows that reasonably well fitting probabilistic models can
be established for circular data by specifying an appropriate distribution family 
such as the von Mises distribution and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable $Y$. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

>>>>

\Abstract{
As circular data appears in many different areas, the development of appropriate statistical 
models which also account for periodicity has become an important but challenging research field. 
For probabilistic modeling of circular data typically the von Mises distribution is employed.
In order to regress its parameters on covariates, a circular distributional tree is proposed as 
an alternative to more commonly-used additive models. The resulting circular trees are 
easy to interpret, can detect non-additive effects, and select covariates and their 
interactions automatically.
By combining an ensemble of circular trees a more stable and regularized circular distributional forest
can be built which allows for modeling smooth effects as well.
%To stabilize and regularize the model an ensemble of distributional trees can be combined 
%yielding a distributional forest which allows for modeling smooth effects as well.

To evaluate the proposed circular trees and forests, a case study on probabilistic 
forecasting of wind direction for two airports in Austria is presented. 
In addition to the tree-based models, alternative common approaches are 
used as a benchmark in order to assess the predictive performances of circular trees 
and forests.

%In order to assess the predictive performance of the proposed circular trees and forests,
%a case study on probabilistic forecasting of wind direction for two airports in Austria is presented
%with alternative commonly used approaches being employed as benchmarks.

%As circular data can be found in many different areas, the development of
%appropriate statistical models accounting, i.a., for periodicity has become a
%challenging and important research field. For probabilistic forecasting
%typically the von Mises distribution is employed in an additive model
%framework. Combining their underlying distributional setup with tree-based
%estimation methods, circular distributional trees will be presented. The resulting
%distributional trees are easy to interpret, can detect non-additive effects,
%and select covariates and their interactions automatically. By combining
%ensembles of circular trees a stabilized circular forest can build, which
%automatically regularizes the estimation and, hence, allows for modeling smooth effects.

%To validate the proposed circular trees and forests, a case study on
%probabilistic forecasting of wind direction will be shown for two airports in
%Austria. In addition to the tree-based models, alternative commonly used
%approaches will be employed to compare their predictive performances to the
%proposed circular trees and forest.

%%%%%%

@@

\begin{figure}[t]
%\centering
%\hspace{-1cm}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \vspace{-1em}
  \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \includegraphics[width=1.2\linewidth]{density_circular.pdf}
\end{minipage}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\caption{Illustration of a distributional model fitted to circular data by 
  maximum likelihood-estimation employing the von Mises distribution. 
  Left panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. 
  Right panel: circular illustration.
  In both panels the empirical histogram (gray) and fitted density (red line) are depicted along 
  with the estimated location parameter (red hand).
  }
  \label{fig:densities} 
  %\includegraphics[width=4cm]{density_vM.pdf}
\end{figure}
%\fixme{update figure 2d to 3d, using data of example tree}

Figure~\ref{fig:densities} depicts a distributional model fitted to circular data
employing the von Mises distribution. While the left panel only serves as an illustration 
of the corresponding  ``linear'' description restricted to an interval of length $2\pi$ but 
consequently not accounting for periodicity the right panel displays an appropriate 
representation of a circular quantity on the unit circle. 
% on observed wind direction data from the airport of Innsbruck.
Looking at the fitted density (red line) and estimated location (red hand) this example 
shows that reasonably well fitting probabilistic models can be established for circular 
data by specifying an appropriate distribution family such as the von Mises distribution
and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable $Y$. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.
